Safe Transferable Regions

### 2016-10-06 paper/type-inference.tex
type inference

\newcommand{\soln}{\eta}

@@

Fig.~\ref{fig:constraint-gen} illustrates this for selected language constructs.
These rules use the same context as the corresponding typing judgment in
Fig.~\ref{fig:fb-staticsem} with symbols $\A$ and $\env$ retaining their meaning.
The algorithm proceeds top-down, analyzes an expression $e$ in a context
$\stdcontext$, and returns a type $\tau$ and a set of constraints $C$
(expressed in the rules as $\exprok {\stdcontext} {e} {\tau} {C}$ ),
indicating that the expression will have a type $\tau$ provided the constraints $C$ hold.

The rules for generating constraints from a method and
class definition first build a context ($\A$) containing a set ($ \rhoenv$) of region variables denoting
regions that are currently live, a map ($\aenv$) mapping type
variables to their bounds, and a constraint formula ($\phicx$)
capturing constraints over live region variables. We use predicate
variables ($\varphi$ and $\varphi_m$) to capture constraints over
variables in $\rhoenv$ denoting the fact that such constraints are yet
to be inferred.

@@

\begin{theorem}
  \label{thm:constraint-generation-sc}
Let $C$ be the set of constraints generated from a program $q$.
Let $\soln$ be any substitution for the region and predicate variables in $q$.
$\soln$ is a solution to $C$ iff $q[\soln]$ is well-typed.
\end{theorem}
\begin{proof}
  We inductively establish a correspondence between derivation trees under
  the static semantics and the constraint-generation trees.

% We do not present a formal proof of the theorem.
The proof essentially requires establishing the following lemma inductively: if
$\exprok {\stdcontext}{e}{\tau}{C}$
and $\soln$ satisfies $C$, then
$\hastyp {\stdcontext}{e[\soln]}{\tau}$,
where $e[\soln]$ indicates $e$ with the substitutions
in the solution $\soln$, as well as analogous lemmas for type well-formedness.
\end{proof}

@@

\paragraph{Constraint Solver.}
Solving the constraints is an iterative process, consisting of the following
steps, and this process fails at any point if it determines that no solution
exists for the set of constraints:

>>>>

\newcommand{\soln}{\eta}
\newcommand{\saturate}[1]{{#1}^*}
\newcommand{\groundc}[1]{\saturate{#1}_g}

@@

Fig.~\ref{fig:constraint-gen} illustrates this for selected language constructs.
These rules use the same context as the corresponding typing judgment in
Fig.~\ref{fig:fb-staticsem}, except that this is generalized to permit the
use of region variables and predicate variables.
Symbols $\A$ and $\env$ retain their meaning, modulo this extension.
The algorithm proceeds top-down, analyzes an expression $e$ in a context
$\stdcontext$, and returns a type $\tau$ and a set of constraints $C$
(expressed in the rules as $\exprok {\stdcontext} {e} {\tau} {C}$ ),
indicating that the expression will have a type $\tau$ provided the constraints $C$ hold.

The rules for generating constraints from a method and
class definition first build a context ($\A$) containing a set ($ \rhoenv$) denoting
regions that are currently live, a map ($\aenv$) mapping type
variables to their bounds, and a constraint formula ($\phicx$)
capturing constraints over live region variables. We use predicate
variables ($\varphi$ and $\varphi_m$) to capture constraints over
variables in $\rhoenv$ denoting the fact that such constraints are yet
to be inferred.

@@

\begin{theorem}
  \label{thm:constraint-generation-sc}
Let $C$ be the set of constraints generated from a program $q$.
Let $\soln$ be any substitution for the region and predicate variables in $q$.
$\soln$ is a solution to $C$ iff $q[\soln]$ is well-typed.
\end{theorem}
\begin{proof}
The correspondence between the static semantics rules and the constraint generation
rules induces a correspondence between derivation trees produced by the
two systems.
% We do not present a formal proof of the theorem.
We can establish the following lemma inductively:
given a derivation tree $\zeta_1$ for
$\exprok {\stdcontext}{e}{\tau}{C}$
and any substitution $\soln$,
we can construct a corresponding \emph{candidate} derivation
tree $\zeta_2$ of $\hastyp {\A[\soln],\env[\soln]}{e[\soln]}{\tau[\soln]}$.
We can show that $\zeta_2$ is a valid derivation tree iff $\soln$ satisfies $C$.
(We make use of analogous lemmas for type well-formedness rules as well.)
\end{proof}

@@

\begin{lemma}
The antecedent of any generated validity constraint is of the form
$\varphi \conj \phictxt$ where $\varphi$ is a predicate variable and
$\phictxt$ is a conjunction of zero or more outlives-constraints.
Let $\pi_1 \outlives \pi_2$ be a conjunct in $\phictxt$.
Then, $\pi_2 \notin \predDeltaMap(\varphi)$.
Furthermore, if $\pi_1 \in \predDeltaMap(\varphi)$, then for
every $\pi_f \in \predDeltaMap(\varphi)$, $\pi_f \outlives \pi_2$ is
a conjunct in $\phictxt$.
\end{lemma}

\begin{proof}
  By induction over the constraint-generation rules.
  Any context $\A = (\rhoenv,\aenv,\phicx)$ generated by the constraint-generation
  process satisfies the invariant that $\phicx$ is of the form $\varphi \conj \phictxt$
  where $\rhoenv \supseteq \predDeltaMap(\varphi)$.
  The only rule that modifies $\phicx$ is the rule for \C{letregion}
  that adds the set of constraints $\pi_f \outlives \pi$ for every $\pi_f \in \rhoenv$
  as conjuncts to $\phicx$.
\end{proof}

\begin{lemma}
  Consider any constraint satisfying the above invariant with a rhs of the
  form $\pi_i \outlives \pi_j$, with both $\pi_i$ and $\pi_j$ region constants.
  (a) If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$,
  then $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ in any solution to the constraints.
  (b) If $\{ \pi_i, \pi_j \} \not \subseteq \predDeltaMap(\varphi)$,
  then $\isvalid{\phicx}{\pi_i \outlives \pi_j}$ in any solution to the constraints.
\end{lemma}

\paragraph{Constraint Solver.}
Our algorithm for solving a set of constraints $C$ consists of the following steps:
\begin{enumerate}
\item We use an iterative fixed point (saturation) algorithm to compute
$\saturate{C}$, the set of all constraints logically implied by $C$.
\item Let $\groundc{C}$ denote the subset of all ground constraints
(\ie, constraints without any region variable or predicate variable)
in $\saturate{C}$. We check if $\groundc{C}$ is valid in the theory of
partial-order. If $\groundc{C}$ is valid, the solution $\soln$ is determined
as below. Otherwise, the constraint solver fails, as no valid solution exists.
\begin{itemize}
\item For any predicate variable $\varphi$, $\predSubstFn(\varphi)$ is defined
to be $\conj \{ \pi_1 \outlives \pi_2 \;|\; \pi_1, \pi_2 \in \regionConstants, \isvalid{\varphi}{\pi_1 \outlives \pi_2} \in \saturate{C} \}$.
\item For any region variable $\rho$,  we are guaranteed that there exists a unique
 $\pi \in \regionConstants$  such that $\saturate{C}$ contains some constraint
with a consequent $\rho = \pi$. We define $\regionSubstFn(\rho)$ to be $\pi$.
\end{itemize}
\end{enumerate}

Solving the constraints is an iterative process, consisting of the following
steps, and this process fails at any point if it determines that no solution
exists for the set of constraints:

### 2016-10-06 paper/typeinference-proofs.tex
do
type inference

Invariants, other points, and to-dos:
\begin{itemize}
  \item use of a set to represent conjunctions
  \item equality is shorthand for two inequalities
  \item unify notation for substitution
  \item substitution required in correctness theorem as well
  \item use of alpha-renaming, and its use to discharge not-in contraints;
also, incorporate this into first stage rather than second (in Fig. 7).
  \item Modify CSP definition to be same as preceding; we can define the $\Delta$s 
    as auxiliary notation
  \item Note that we can also clarify that the well-formedness constraints are
    syntactic sugar from basic set constraints
  \item Every predicate variable has a single domain-constraint.
  \item Prove: Progress: Every region-variable will be unified with some constant.
  \item Notation: should we use a different notation to distinguish a validity constraint
    from actual entailment?
  \item Proof by relating proof-trees with constraint-generation trees; each can
    be constructed from the other; inductive lemma is as earlier
\end{itemize}

\begin{lemma}
The antecedent of any generated validity constraint is of the form
$\varphi \conj \phictxt$ where $\varphi$ is a predicate variable and
$\phictxt$ is a conjunction of zero or more outlives-constraints.
Let $\pi_1 \outlives \pi_2$ be a conjunct in $\phictxt$.
Then, $\pi_2 \notin \predDeltaMap(\varphi)$.
Furthermore, if $\pi_1 \in \predDeltaMap(\varphi)$, then for
every $\pi_f \in \predDeltaMap(\varphi)$, $\pi_f \outlives \pi_2$ is
a conjunct in $\phictxt$.
\end{lemma}

\begin{proof}
  By induction over the constraint-generation rules.
  Any context $\A = (\rhoenv,\aenv,\phicx)$ generated by the constraint-generation
  process satisfies the invariant that $\phicx$ is of the form $\varphi \conj \phictxt$
  where $\rhoenv \supseteq \predDeltaMap(\varphi)$.
  The only rule that modifies $\phicx$ is the rule for \C{letregion}
  that adds the set of constraints $\pi_f \outlives \pi$ for every $\pi_f \in \rhoenv$
  as conjuncts to $\phicx$.
\end{proof}

\begin{lemma}
  Consider any constraint satisfying the above invariant with a rhs of the
  form $\pi_i \outlives \pi_j$, with both $\pi_i$ and $\pi_j$ region constants.
  (a) If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$,
  then $\isvalid{\varphi}{\pi_i \outlives \pi_j}$ in any solution to the constraints.
  (b) If $\{ \pi_i, \pi_j \} \not \subseteq \predDeltaMap(\varphi)$,
  then $\isvalid{\phicx}{\pi_i \outlives \pi_j}$ in any solution to the constraints.
\end{lemma}

>>>>

Invariants, other points, and to-dos:
\begin{itemize}
  \item discuss computation of least-solution to constraints?
  \item use of a set to represent conjunctions
  \item equality is shorthand for two inequalities
  \item unify notation for substitution
  \item substitution required in correctness theorem as well
  \item use of alpha-renaming, and its use to discharge not-in contraints;
also, incorporate this into first stage rather than second (in Fig. 7).
  \item Modify CSP definition to be same as preceding; we can define the $\Delta$s 
    as auxiliary notation
  \item Note that we can also clarify that the well-formedness constraints are
    syntactic sugar from basic set constraints
  \item Every predicate variable has a single domain-constraint.
  \item Prove: Progress: Every region-variable will be unified with some constant.
  \item Notation: should we use a different notation to distinguish a validity constraint
    from actual entailment?
  \item Proof by relating proof-trees with constraint-generation trees; each can
    be constructed from the other; inductive lemma is as earlier
\end{itemize}
